Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-19 ind_files/03_BD_processes.tex
Add some references
Minor phrasing changes after proofreading

Let us now consider a slightly more complicated scenario. Assume that our population is \emph{not} composed of identical organisms, but instead can contain up to $m$ different kinds of organisms - For example, individuals may come in one of $m$ colors, or a gene may have $m$ different alleles. The specific interpretation of the different variants is irrelevant to our formalism, and I, therefore, refer to each distinct variant of an organism simply as a `type'. Unlike many classic stochastic formulations in evolutionary theor~\citep{fisher_genetical_1930,wright_evolution_1931,moran_random_1958,crow_introduction_1970, lande_natural_1976, kimura_probability_1974}, I do not assume a fixed (effective) population size and instead allow the total population size to fluctuate naturally over time.

@@

\section{The system-size expansion}
As before, we now assume we can find a system size parameter $K > 0$ such that we can make the substitutions
\begin{align*}
\mathbf{x} &= \frac{\mathbf{n}}{K}\\
b^{(K)}_i(\mathbf{x}) &= \frac{1}{K}b_i(\mathbf{n})\\
d^{(K)}_i(\mathbf{x}) &= \frac{1}{K}d_i(\mathbf{n})
\end{align*}
and define new step operators $\Delta_{i}^{\pm}$ by their action on any real-valued function $f(\mathbf{x},t)$ as
\begin{equation}
\label{nD_step_operators_rescaled}
\Delta_{i}^{\pm}f([x_1,\ldots,x_m],t) = f([x_1,\ldots,x_i \pm \frac{1}{K}, \ldots x_m],t)
\end{equation}
In terms of these new variables, \eqref{nD_M_eqn} becomes
\begin{equation}
\label{nd_M_eqn_rescaled}
\frac{\partial P}{\partial t}(\mathbf{x},t) = K\sum\limits_{j=1}^{m}\left[(\Delta_j^{-}-1)b^{(K)}_j(\mathbf{x})P(\mathbf{x},t) + (\Delta_j^{+}-1)d^{(K)}_j(\mathbf{x})P(\mathbf{x},t)\right]
\end{equation}
If $K$ is large, we can once again Taylor expand the action of the step operators as
\begin{equation*}
f([x_1,\ldots,x_i \pm \frac{1}{K}, \ldots x_m],t) = f(\mathbf{x},t) \pm \frac{1}{K}\frac{\partial f}{\partial x_i}(\mathbf{x},t) + \frac{1}{2K^2}\frac{\partial^2f}{\partial x_i^2}(\mathbf{x},t) + \mathcal{O}(K^{-3})
\end{equation*}
which, after substituting into \eqref{nd_M_eqn_rescaled}, yields the equation
\begin{equation}
\label{nD_FPE}
\setlength{\fboxsep}{2\fboxsep}\boxed{\frac{\partial P}{\partial t}(\mathbf{x},t) = \sum\limits_{j=1}^{m}\left[-\frac{\partial}{\partial x_j}\{A_j^{-}(\mathbf{x})P(\mathbf{x},t)\} + \frac{1}{2K}\frac{\partial^2}{\partial x_j^2}\{A_j^{+}(\mathbf{x})P(\mathbf{x},t)\}\right]}
\end{equation}
where
\begin{equation*}
A_{i}^{\pm}(\mathbf{x}) = b^{(K)}_i(\mathbf{x})\pm d^{(K)}_i(\mathbf{x})
\end{equation*}
Equation \eqref{nD_FPE} is an $m$-dimensional Fokker-Planck equation, and corresponds to the $m$-dimensional It\^o process
\begin{equation}
\label{nD_Ito_SDE}
d\mathbf{X}_{t} = \mathbf{A^-}(\mathbf{X}_t)dt + \frac{1}{\sqrt{K}}\mathbf{D}(\mathbf{X}_t)d\mathbf{W}_t
\end{equation}
where $\mathbf{A^-}(\mathbf{X}_t)$ is the $m$ dimensional `drift vector' with $i$\textsuperscript{th} element $ = A^{-}_{i}(\mathbf{X}_t)$. $\mathbf{D}(\mathbf{X}_t)$ is the $m \times m$ `diffusion matrix' with $ij$th element $\left(\mathbf{D}(\mathbf{X}_t)\right)_{ij} = \delta_{ij}\left(A^{+}_{i}A^{+}_{j}\right)^{\frac{1}{4}}$, where $\delta_{ij}$ is the Kronecker delta symbol, defined by
\begin{equation*}
\delta_{ij} = 
\begin{cases}
1 & i=j\\
0 & i\neq j
\end{cases}
\end{equation*}
Finally, $\mathbf{W}_t$ is the $m$-dimensional Wiener process and can be thought of as a vector of independent one-dimensional Wiener processes (which have been defined in \ref{intro_SDE}). This is the `mesoscopic' description of our process.

@@

We assume that the birth and death rate functions have the functional form
\begin{equation}
\label{nD_functional_forms_for_replicator}
\begin{aligned}
b^{(K)}_i(\mathbf{x}) &= x_ib^{\textrm{(ind)}}_{i}(\mathbf{x}) + \mu Q_i(\mathbf{x})\\
d^{(K)}_i(\mathbf{x}) &= x_id^{\textrm{(ind)}}_i(\mathbf{x})
\end{aligned}
\end{equation}
where $b^{\textrm{(ind)}}_{i}(\mathbf{x})$ and $d^{\textrm{(ind)}}_{i}(\mathbf{x})$ are non-negative functions that respectively describe the per-capita birth and death rate of type $i$ individuals, $\mu \geq 0$ is a constant describing the mutation rate in the population, and $Q_i(\mathbf{x})$ is a non-negative function that describes the additional birth rate of type $i$ individuals due to mutations in the population $\mathbf{x}$ that cannot be captured in the per-capita birth rate\footnote{When $x_i = 0$, \emph{i.e.} there are no type $i$ individuals in the population, individuals of type $i$ may still be born through mutations during births of the other types. This cannot be captured in $b^{\textrm{(ind)}}_{i}(\mathbf{x})$ because the term $x_ib^{\textrm{(ind)}}_{i}(\mathbf{x})$ vanishes when $x_i = 0$. Note that no analogous problem exists for the death rate, since the death rate of type $i$ individuals must be 0 when $x_i$ is 0 to ensure that we never have negative population densities.}. Our assumptions of the functional forms \eqref{nD_functional_forms_for_replicator} thus amount to saying that birth and death rates can be separated into mutational and non-mutational components, and furthermore that the density dependence of the birth and death rates due to non-mutational effects is in a form that allows us to write down per-capita birth and death rates for each type. We define the \emph{Malthusian fitness} of the $i\textsuperscript{th}$ type as $w_i(\mathbf{x}) \coloneqq b^{\textrm{(ind)}}_{i}(\mathbf{x}) - d^{\textrm{(ind)}}_{i}(\mathbf{x})$, and the \emph{per-capita turnover rate} of the $i\textsuperscript{th}$ type as $\tau_i(\mathbf{x}) = b^{\textrm{(ind)}}_{i}(\mathbf{x}) + d^{\textrm{(ind)}}_{i}(\mathbf{x})$.  The quantity $w_i(\mathbf{x})$ describes the per-capita growth rate of type $i$ individuals in a population $\mathbf{x}$ discounting mutation. Ecologists often denote this quantity by the symbol $r_i$ and simply call it the (exponential) growth rate of type $i$, but I will stick to $w_i$ and `fitness' here. It is notable that both $w_i$ and $\tau_i$ depend on the state of the population as a whole (\textit{i.e.} $\mathbf{x}$) and not just on the density of the focal type. Thus, in general, both the fitness and the turnover rate in our model are frequency-dependent.

\section{Statistical measures for population-level quantities}\label{sec_stat_measures}
Given a state $\mathbf{x}(t)$ that describes our population at time $t$, we can compute the total (scaled) population size and the frequency of each type in the population as:
\begin{equation}
\label{nD_tot_pop_and_prop_inds_defn}
\begin{aligned}
N_{K}(t) &\coloneqq \sum\limits_{i=1}^{m}x_i(t) = \frac{1}{K}\sum\limits_{i=1}^{m}v_i(t)\\
p_i(t) &\coloneqq \frac{x_i(t)}{N_{K}(t)}
\end{aligned}
\end{equation}
These definitions are important because evolution is often measured as changes in type frequency (and \emph{not} necessarily changes in type density). Note that type frequency is a population level. Indeed, even though the causes of evolution are generally at the individual level, we are very often interested in quantities \emph{described} at a population level, such as the mean fitness or mean phenotype in the population~\citep{bourrat_evolution_2019}. Furthermore, the relevant quantities at the individual level, such as individual fitness or phenotype, are typically equal for all individuals of the same type (in some sense this is our basis for defining different types in the first place), and we, therefore, call them `type-level' quantities below.

Let $f$ be any type-level quantity with (possibly time-dependent) value $f_i$ for the $i\textsuperscript{th}$ type. For example, if each type is a phenotype for a trait such as height, which can be assigned a numerical value, then setting $f_i = \textit{value of $i\textsuperscript{th}$ phenotype}$ gives us the mean trait value in the population. To facilitate the description of such quantities, we define the statistical mean value of any quantity $f$ in the population as
\begin{equation}
\label{nD_mean}
\overline{f}(t) \coloneqq \sum\limits_{i=1}^{m}f_ip_{i}
\end{equation}
the statistical covariance of two such quantities $f$ and $g$ as
\begin{equation}
\label{nD_cov}
\textrm{Cov}(f,g) \coloneqq \overline{fg} - \overline{f}\overline{g}
\end{equation}
and the statistical variance of a quantity $f$ as $\sigma^2_f \coloneqq \textrm{Cov}(f,f)$.

It is important to recognize that these quantities are distinct from and independent of the \emph{probabilistic} expectation, variance, and covariance obtained by integrating over realizations in the underlying probability space. Indeed, for finite populations, the statistical mean, statistical variance, and statistical covariance are all themselves stochastic processes: For each instant of time, these population-level quantities are a random variable (\emph{i.e.} a \emph{function} and not just a number) depending on $\mathbf{p}$, the (random) vector of type frequencies in the population. For infinite populations, the statistical mean, variance, and covariance are entirely deterministic time-dependent quantities that simply describe how $f$ is distributed across the population. Failure to clearly make this distinction between statistical operations and probabilistic operations has led to much confusion with regard to the infinite population Price equation~\citep{van_veelen_use_2005}, which is entirely deterministic.

@@

In appendix \ref{App_density_to_freq}, I use It\^{o}'s formula to derive a general stochastic equation for the frequencies of each type in the population. Letting $\overline{w} = \sum w_ip_i$ and $\overline{\tau} = \sum \tau_i p_i$ be the average population fitness and turnover respectively, I show in appendix \ref{App_density_to_freq} that the frequency of the $i\textsuperscript{th}$ type in the population $\mathbf{x}(t)$ changes according to the equation:
\begin{equation}
\label{nD_eqn_for_frequencies}
\begin{aligned}
dp_i(t) &= \left[(w_i(\mathbf{x}) - \overline{w})p_i + \mu\left\{Q_i(\mathbf{p}) - p_i\left(\sum\limits_{j=1}^{m}Q_j(\mathbf{p})\right)\right\}\right]dt\\
&- \frac{1}{K}\frac{1}{N_{K}(t)}\left[(\tau_i(\mathbf{x}) - \overline{\tau})p_i + \mu\left\{Q_i(\mathbf{p}) - p_i\left(\sum\limits_{j=1}^{m}Q_j(\mathbf{p})\right)\right\}\right]dt\\
&+ \frac{1}{\sqrt{K}}\frac{1}{N_{K}(t)}\left[\left(A^{+}_{i}\right)^{1/2}dW^{(i)}_t - p_i\sum\limits_{j=1}^{m}\left(A^{+}_{j}\right)^{1/2}dW^{(j)}_t\right]
\end{aligned}
\end{equation}
where $W^{(1)}_t,W^{(2)}_t, \ldots, W^{(m)}_t$ are $m$ independent one-dimensional standard Brownian motion processes and I have used the notation $Q_i(\mathbf{p}) = Q_i(\mathbf{x})/N_K(t)$ for notational clarity. I will show below that the first term in this expression describes directional changes in the population composition due to `classical' evolutionary forces such as selection and mutation that occur in deterministic infinite population models. The second term is an additional directional force on population composition that is only seen in finite populations and can be thought of as a biasing `selection' for reduced turnover rate due to an effect similar to gambler's ruin in probability theory. The consequences of this term, as well as connections with previous studies, are discussed in detail in part \ref{part_summary}. Finally, the last term of equation \eqref{nD_eqn_for_frequencies} describes non-directional stochastic effects due to fluctuations and has a `spreading effect'.

@@

This is a Price equation for quantities $f_i$ which can vary over time~\citep{lion_theoretical_2018}. To obtain the more familiar Price equation seen in textbooks, we can consider time-independent $f_i$, \emph{i.e.} situations in which each $f_i$ is constant over time, and thus changes in $\overline{f}$ are purely due to changes in the composition of the population. For such quantities, we have $\frac{\partial f_i}{\partial t} = 0 \ \forall \ i$ and thus obtain the dynamic version of the famous Price equation~\citep{page_unifying_2002,lion_theoretical_2018}:

>>>>

Let us now consider the situation we are actually interested in. Assume that our population is \emph{not} composed of identical organisms, but instead can contain up to $m$ different kinds of organisms - For example, individuals may come in one of $m$ colors, or a gene may have $m$ different alleles. The specific interpretation of the different variants is irrelevant to our formalism, and I, therefore, refer to each distinct variant of an organism simply as a `type'. Unlike many classic stochastic formulations in evolutionary theory~\citep{fisher_genetical_1930,wright_evolution_1931,moran_random_1958, kimura_problems_1957, kimura_diffusion_1964, kimura_number_1964, crow_introduction_1970, lande_natural_1976}, I do not assume a fixed (effective) population size and instead allow the total population size to fluctuate naturally over time.

@@

\section{The system-size expansion}
As before, we now assume we can find a system size parameter $K > 0$ such that we can make the substitutions
\begin{align*}
\mathbf{x} &= \frac{\mathbf{n}}{K}\\
b^{(K)}_i(\mathbf{x}) &= \frac{1}{K}b_i(\mathbf{n})\\
d^{(K)}_i(\mathbf{x}) &= \frac{1}{K}d_i(\mathbf{n})
\end{align*}
and define new step operators $\Delta_{i}^{\pm}$ by their action on any real-valued function $f(\mathbf{x},t)$ as
\begin{equation}
\label{nD_step_operators_rescaled}
\Delta_{i}^{\pm}f([x_1,\ldots,x_m],t) = f([x_1,\ldots,x_i \pm \frac{1}{K}, \ldots x_m],t)
\end{equation}
In terms of these new variables, \eqref{nD_M_eqn} becomes
\begin{equation}
\label{nd_M_eqn_rescaled}
\frac{\partial P}{\partial t}(\mathbf{x},t) = K\sum\limits_{j=1}^{m}\left[(\Delta_j^{-}-1)b^{(K)}_j(\mathbf{x})P(\mathbf{x},t) + (\Delta_j^{+}-1)d^{(K)}_j(\mathbf{x})P(\mathbf{x},t)\right]
\end{equation}
If $K$ is large, we can once again Taylor expand the action of the step operators as
\begin{equation*}
f([x_1,\ldots,x_i \pm \frac{1}{K}, \ldots x_m],t) = f(\mathbf{x},t) \pm \frac{1}{K}\frac{\partial f}{\partial x_i}(\mathbf{x},t) + \frac{1}{2K^2}\frac{\partial^2f}{\partial x_i^2}(\mathbf{x},t) + \mathcal{O}(K^{-3})
\end{equation*}
which, after substituting into \eqref{nd_M_eqn_rescaled}, yields the equation
\begin{equation}
\label{nD_FPE}
\setlength{\fboxsep}{2\fboxsep}\boxed{\frac{\partial P}{\partial t}(\mathbf{x},t) = \sum\limits_{j=1}^{m}\left[-\frac{\partial}{\partial x_j}\{A_j^{-}(\mathbf{x})P(\mathbf{x},t)\} + \frac{1}{2K}\frac{\partial^2}{\partial x_j^2}\{A_j^{+}(\mathbf{x})P(\mathbf{x},t)\}\right]}
\end{equation}
where
\begin{equation*}
A_{i}^{\pm}(\mathbf{x}) = b^{(K)}_i(\mathbf{x})\pm d^{(K)}_i(\mathbf{x})
\end{equation*}
Equation \eqref{nD_FPE} is an $m$-dimensional Fokker-Planck equation, and corresponds to the $m$-dimensional It\^o process
\begin{equation}
\label{nD_Ito_SDE}
d\mathbf{X}_{t} = \mathbf{A^-}(\mathbf{X}_t)dt + \frac{1}{\sqrt{K}}\mathbf{D}(\mathbf{X}_t)d\mathbf{W}_t
\end{equation}
where $\mathbf{A^-}(\mathbf{X}_t)$ is an $m$-dimensional vector with $i$\textsuperscript{th} element $ = A^{-}_{i}(\mathbf{X}_t)$. $\mathbf{D}(\mathbf{X}_t)$ is an $m \times m$ matrix with $ij$th element $\left(\mathbf{D}(\mathbf{X}_t)\right)_{ij} = \delta_{ij}\left(A^{+}_{i}A^{+}_{j}\right)^{\frac{1}{4}}$, where $\delta_{ij}$ is the Kronecker delta symbol, defined by
\begin{equation*}
\delta_{ij} = 
\begin{cases}
1 & i=j\\
0 & i\neq j
\end{cases}
\end{equation*}
Finally, $\mathbf{W}_t$ is the $m$-dimensional Wiener process and can be thought of as a vector of independent one-dimensional Wiener processes (which have been defined in \ref{intro_SDE}). This is the `mesoscopic' description of our process.

@@

I assume that the birth and death rate functions have the functional form
\begin{equation}
\label{nD_functional_forms_for_replicator}
\begin{aligned}
b^{(K)}_i(\mathbf{x}) &= x_ib^{\textrm{(ind)}}_{i}(\mathbf{x}) + \mu Q_i(\mathbf{x})\\
d^{(K)}_i(\mathbf{x}) &= x_id^{\textrm{(ind)}}_i(\mathbf{x})
\end{aligned}
\end{equation}
where $b^{\textrm{(ind)}}_{i}(\mathbf{x})$ and $d^{\textrm{(ind)}}_{i}(\mathbf{x})$ are non-negative functions that respectively describe the per-capita birth and death rate of type $i$ individuals, $\mu \geq 0$ is a constant describing the mutation rate in the population, and $Q_i(\mathbf{x})$ is a non-negative function that describes the additional birth rate of type $i$ individuals due to mutations in the population $\mathbf{x}$ that cannot be captured in the per-capita birth rate\footnote{When $x_i = 0$, \emph{i.e.} there are no type $i$ individuals in the population, individuals of type $i$ may still be born through mutations during births of the other types. This cannot be captured in $b^{\textrm{(ind)}}_{i}(\mathbf{x})$ because the term $x_ib^{\textrm{(ind)}}_{i}(\mathbf{x})$ vanishes when $x_i = 0$. Note that no analogous problem exists for the death rate, since the death rate of type $i$ individuals must be 0 when $x_i$ is 0 to ensure that we never have negative population densities.}. My assumptions on the functional forms \eqref{nD_functional_forms_for_replicator} thus amount to saying that birth and death rates can be separated into mutational and non-mutational components, and furthermore that the density dependence of the birth and death rates due to non-mutational effects is in a form that allows us to write down per-capita birth and death rates for each type. Let us define the \emph{Malthusian fitness} of the $i\textsuperscript{th}$ type as $w_i(\mathbf{x}) \coloneqq b^{\textrm{(ind)}}_{i}(\mathbf{x}) - d^{\textrm{(ind)}}_{i}(\mathbf{x})$, and the \emph{per-capita turnover rate} of the $i\textsuperscript{th}$ type as $\tau_i(\mathbf{x}) = b^{\textrm{(ind)}}_{i}(\mathbf{x}) + d^{\textrm{(ind)}}_{i}(\mathbf{x})$.  The quantity $w_i(\mathbf{x})$ describes the per-capita growth rate of type $i$ individuals in a population $\mathbf{x}$ discounting mutation. Ecologists often denote this quantity by the symbol $r_i$ and simply call it the (exponential) growth rate of type $i$, but I will stick to $w_i$ and `fitness' here. $\tau_i$ is a measure of the number of events (birth events + death events) that a type $i$ individual experiences in a given time interval --- Individuals with higher turnover rates experience more events (on average) than those with lower turnover rates. This can be thought of as a measure of the `pace of life'. I briefly note that the quantity $\tau_i$ has also been called the `variability in the reproductive output' in the literature~\citep{gillespie_natural_1974}. It is notable that both $w_i$ and $\tau_i$ depend on the state of the population as a whole (\textit{i.e.} $\mathbf{x}$) and not just on the density of the focal type. Thus, in general, the fitness and the turnover rate in our model are both frequency and density-dependent.

\section{Statistical measures for population-level quantities}\label{sec_stat_measures}
Given a state $\mathbf{x}(t)$ that describes our population at time $t$, we can compute the total (scaled) population size and the frequency of each type in the population as:
\begin{equation}
\label{nD_tot_pop_and_prop_inds_defn}
\begin{aligned}
N_{K}(t) &\coloneqq \sum\limits_{i=1}^{m}x_i(t) = \frac{1}{K}\sum\limits_{i=1}^{m}v_i(t)\\
p_i(t) &\coloneqq \frac{x_i(t)}{N_{K}(t)}
\end{aligned}
\end{equation}
These definitions are important because evolution is often defined via changes in type frequency (and \emph{not} necessarily changes in type density).

Even though the causes of evolution are generally at the individual level, we are very often interested in quantities \emph{described} at a population level, such as the mean fitness or mean phenotype in the population~\citep{bourrat_evolution_2019}. Furthermore, the relevant quantities at the individual level, such as individual fitness or phenotype, are typically equal for all individuals of the same type (in some sense this is our basis for defining different types in the first place). I use the term `type-level quantities' henceforth to refer to such quantities that are equal for all individuals that are of the same type.

Let $f$ be any type-level quantity with (possibly time-dependent) value $f_i$ for the $i\textsuperscript{th}$ type. For example, if each type is a phenotype for a trait such as height, which can be assigned a numerical value, then setting $f_i = \textit{value of $i\textsuperscript{th}$ phenotype}$ gives us the mean trait value in the population. To facilitate the description of such quantities, I define the statistical mean value of any quantity $f$ in the population as
\begin{equation}
\label{nD_mean}
\overline{f}(t) \coloneqq \sum\limits_{i=1}^{m}f_ip_{i}
\end{equation}
the statistical covariance of two such quantities $f$ and $g$ as
\begin{equation}
\label{nD_cov}
\textrm{Cov}(f,g) \coloneqq \overline{fg} - \overline{f}\overline{g}
\end{equation}
and the statistical variance of a quantity $f$ as $\sigma^2_f \coloneqq \textrm{Cov}(f,f)$.

It is important to recognize that these quantities are distinct from and independent of the \emph{probabilistic} expectation, variance, and covariance obtained by integrating over realizations in the underlying probability space. Indeed, for finite populations, the statistical mean, statistical variance, and statistical covariance are all themselves stochastic processes: For each instant of time, these population-level quantities are a random variable (\emph{i.e.} a \emph{function} and not just a number) depending on $\mathbf{p}$, the (random) vector of type frequencies in the population. For infinite populations, the statistical mean, variance, and covariance are entirely deterministic time-dependent quantities that simply describe how $f$ is distributed across the population. Failure to clearly make this distinction between statistical operations and probabilistic operations has led to much confusion with regard to the infinite population Price equation, which is entirely deterministic and does not incorporate features of finite populations such as drift in its original formulation~\citep{frank_price_1997,van_veelen_use_2005,frank_natural_2012}.

@@

In appendix \ref{App_density_to_freq}, I use a multivariate version of It\^{o}'s formula to derive a general stochastic equation for the frequencies of each type in the population. Letting $\overline{w} = \sum w_ip_i$ and $\overline{\tau} = \sum \tau_i p_i$ be the average population fitness and turnover respectively, I show in appendix \ref{App_density_to_freq} that the frequency of the $i\textsuperscript{th}$ type in the population $\mathbf{x}(t)$ changes according to the equation:
\begin{equation}
\label{nD_eqn_for_frequencies}
\begin{aligned}
dp_i(t) &= \left[(w_i(\mathbf{x}) - \overline{w})p_i + \mu\left\{Q_i(\mathbf{p}) - p_i\left(\sum\limits_{j=1}^{m}Q_j(\mathbf{p})\right)\right\}\right]dt\\
&- \frac{1}{K}\frac{1}{N_{K}(t)}\left[(\tau_i(\mathbf{x}) - \overline{\tau})p_i + \mu\left\{Q_i(\mathbf{p}) - p_i\left(\sum\limits_{j=1}^{m}Q_j(\mathbf{p})\right)\right\}\right]dt\\
&+ \frac{1}{\sqrt{K}}\frac{1}{N_{K}(t)}\left[\left(A^{+}_{i}\right)^{1/2}dW^{(i)}_t - p_i\sum\limits_{j=1}^{m}\left(A^{+}_{j}\right)^{1/2}dW^{(j)}_t\right]
\end{aligned}
\end{equation}
where $W^{(1)}_t,W^{(2)}_t, \ldots, W^{(m)}_t$ are $m$ independent one-dimensional standard Brownian motion processes and I have used the notation $Q_i(\mathbf{p}) = Q_i(\mathbf{x})/N_K(t)$ for notational clarity. I will show below that the first term in this expression describes directional changes in the population composition due to `classical' evolutionary forces such as selection and mutation that occur in deterministic infinite population models. The second term is an additional directional force on population composition that is only seen in finite populations and can be thought of as a biasing `selection' for reduced turnover rate due to an effect similar to gambler's ruin in probability theory. The consequences of this term, as well as connections with previous studies, are discussed in detail in part \ref{part_summary}. Finally, the last term of equation \eqref{nD_eqn_for_frequencies} describes non-directional stochastic effects due to fluctuations and has a `spreading effect'.

@@

This is a Price equation for quantities $f_i$ which can vary over time~\citep{lion_theoretical_2018,day_price_2020}. To obtain the more familiar Price equation seen in textbooks, we can consider time-independent $f_i$, \emph{i.e.} situations in which each $f_i$ is constant over time, and thus changes in $\overline{f}$ are purely due to changes in the composition of the population. For such quantities, we have $\frac{\partial f_i}{\partial t} = 0 \ \forall \ i$ and thus obtain the dynamic version of the famous Price equation~\citep{page_unifying_2002,lion_theoretical_2018}:

### 2023-03-19 ind_files/04_field_eqns.tex
Minor phrasing changes after proofreading

\section{The infinite population limit}
Taking $K \to \infty$ in equation \eqref{functional_langevin} then yields a PDE:
\begin{equation}
\label{deterministic_traj}
\frac{\partial \psi}{\partial t}(x,t) = \mathcal{A}^{-}\left(x|\psi\right) = b_K(x|\psi)- d_K(x|\psi)
\end{equation}
where I have used a different symbol $\psi$ simply to highlight that $\psi(x,t)$ as the solution to equation \eqref{deterministic_traj} is a deterministic function, whereas $\phi(x,t)$ as defined in equation \eqref{functional_langevin} is really a stochastic process $\{\phi^{(t)}\}_{t\geq0}$. Equation \eqref{deterministic_traj} simply says that in the absence of stochasticity, the change in the density of individuals with trait values $x$ is given by the difference between the birth and death rates of these individuals in the population. Models of this form are precisely the `PDE models' discussed in studies of Adaptive Diversification~\citep{doebeli_adaptive_2011}. Equation \eqref{deterministic_traj} is also the basic equation of `oligomorphic dynamics'~\citep{sasaki_oligomorphic_2011, lion_multimorph_2022} if one assumes the population is composed of a small number of `morphs', \emph{i.e.} $\psi(x,t) = \sum\limits_{k=1}^{S} n_{k}\psi_k(x,t)$, where $\psi_k(x,t)$ is the phenotypic distribution of the $k$th morph (often assumed a normal distribution with narrow variance) and $S$ is the number of distinct morphs in the population. Models of the form \eqref{deterministic_traj} are also used to study intraspecific trait variation in community ecology~\citep{nordbotten_dynamics_2020}. A prominent recent example is the `trait space equations' of~\citep{wickman_theoretical_2022} in their framework for eco-evolutionary community dynamics.

>>>>

\section{The infinite population limit}
Taking $K \to \infty$ in equation \eqref{functional_langevin} then yields a PDE:
\begin{equation}
\label{deterministic_traj}
\frac{\partial \psi}{\partial t}(x,t) = \mathcal{A}^{-}\left(x|\psi\right) = b_K(x|\psi)- d_K(x|\psi)
\end{equation}
where I have used a different symbol $\psi$ simply to highlight that $\psi(x,t)$ as the solution to equation \eqref{deterministic_traj} is a deterministic function, whereas $\phi(x,t)$ as defined in equation \eqref{functional_langevin} is really a stochastic process $\{\phi^{(t)}\}_{t\geq0}$. Equation \eqref{deterministic_traj} simply says that in the absence of stochasticity, the change in the density of individuals with trait values $x$ is given by the difference between the birth and death rates of these individuals in the population. Models of this form are precisely the `PDE models' discussed in studies of Adaptive Diversification~\citep{doebeli_adaptive_2011}. Equation \eqref{deterministic_traj} is also the basic equation of `oligomorphic dynamics'~\citep{sasaki_oligomorphic_2011, lion_multimorph_2022} if one assumes the population is composed of a small number of `morphs', \emph{i.e.} $\psi(x,t) = \sum\limits_{k=1}^{S} n_{k}\psi_k(x,t)$, where $\psi_k(x,t)$ is the phenotypic distribution of the $k$th morph (often assumed a normal distribution with narrow variance) and $S$ is the number of distinct morphs in the population. Models of the form \eqref{deterministic_traj} are also used to study intraspecific trait variation in community ecology~\citep{nordbotten_dynamics_2020}. A prominent recent example is~\cite{wickman_theoretical_2022}'s `trait space equations' in their framework for eco-evolutionary community dynamics.
